Dependent Types and Martin-L\"of Type Theory: An Introduction

### 2016-07-25 paper.tex
more

\section{Introduction}
We present an introduction to dependent types and (Martin-L\"of) type theory.
At its core, type theory seeks to serve as a possible foundation of mathematics,
which itself is based on a different logical foundation. This proof-relevant
form of mathematics is the basis in which types can be used to model logical
propositions and is the core idea of type theory. As such, not only does type
theory have a viable application in mathematics but also in computer science,
providing for both a richer world of types and expressibility.

The study of type theory goes hand in hand with the study of dependent types,
which is  why we first take the time to do familiarize ourselves with them in
Section 2. We then continue with the foundational concepts of type theory of
Per Martin-L\"of \cite{}. In this section, while reading through the examples
we have provided, we recommend using a proof assistant, such as Agda. In our
examples, we refer to Agda notation.

@@

\begin{enumerate}
\item terms depending on terms
\item terms depending on types
\item types depending on types
\item types depending on terms
\end{enumerate}
These dependecies, however, should not be confused with an actual dependent
type. In fact, only (2) and (4) correspond to expressions that are typed
(necessarily) with a dependent type (we explain later why (1) and (3) are
not typed with dependent types).

@@

\subsection*{Simply-Typed (STLC)}
\subsubsection*{Role: Base System}
The STLC is not a dependently typed system. In fact, the term {\em dependent type}
is drawn in constrast with the term {\em simple type}, where the STLC derives
its namesake. A simple type is either a {\em primitive type}, which is the set
of all types ``pre-defined'' (e.g. lists, numbers, booleans), or a
{\em function type} between other simple types. Thus, simple types have the
following grammar:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
t || t → t
\end{code}
\end{minipage}
\end{center}
The terms of the STLC are the terms of the \textit{untyped \(\lambda\)-calculus},
with the exception of terms that use \textit{self-application} (i.e., Y
combinator, Omega, etc.). An example term, annotated with its type, is generally
written as follows:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
example : ℕ → ℕ
example = λ (z : ℕ) → z
\end{code}
\end{minipage}
\end{center}
Or, equivalently,
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
example = λ z → z
\end{code}
\end{minipage}
\end{center}
whenever the type of the argument (viz. {\tt z}) is easily inferable.

This leaves the STLC rather small and limited but has the benefit of removing
all \textit{infinite} calculations from the system \cite{}. Thus, the STLC is
not \textit{Turing complete} and is thus the appropriate foundation for
deriving a type theoretic system. This is because the STLC lends itself quite
seamlessly as component of a larger, more expressive system. This flexibility
is what we exploit to derive \(\lambda C\). To do this, we first extend the STLC
with more expressive types through simple extentions on its grammar for simple
types.

%% as the only form of dependency
%% present in the STLC is (1), terms depending on terms (i.e. {\em term level}
%% \(\lambda\)-abstraction).

The key difference between the STLC and \(\lambda C\) are the places in which
terms and types are allowed to be. In the STLC, terms and types are entirely
seperate entities in that terms only exist on {\em term level} and types
on {\em type level}. Thus, as we mentioned earlier, this puts a constraint
on the system to only allow \(\lambda\)-abstraction to happen at term level
(viz. definition of simple type). For example, the \textit{constant function}
over types \(\mathbb{N}\) and \(\mathbb{B}\) in the STLC is defined:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
const : ℕ → 𝔹 → ℕ
const = λ n b → n
\end{code}
\end{minipage}
\end{center}
Given that {\tt const} is of type {\tt ℕ → 𝔹 → ℕ}, the term that inhabits this
type is {\tt λ n b → n}. In this example, we can see that terms and types are
treated differently and are only related in terms of {\em inhabitance}. This
relation is reason for the limited expressivity of the STLC and is also the
method in which the STLC can be extended into \(\lambda C\).

@@

Another name for \(\lambda2\) is System F \cite{}. For our purposes, we treat
\(\lambda2\) as simply an extension of the STLC. This system is adds
{\em polymorphic} terms to the STLC (viz. Dependency (2)). For example,
let's take the definition of {\tt const} above and change it to define
{\tt polyConst}:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
polyConst : (A B : Set) → A → B → A
polyConst A B = λ a b → a
-- Works for any type A and B
\end{code}
\end{minipage}
\end{center}

Unlike {\tt const}, {\tt polyConst} takes four arguments, where the first two
are types. These two types (viz {\tt A} and {\tt B}) are the corresponding
types of the third and fourth arguments of {\tt polyConst}. To clarify, we can
instantiate {\tt polyConst} with \(\mathbb{N}\) and \(\mathbb{B}\) resulting
in a function of type {\tt ℕ → 𝔹 → ℕ}, giving us

$${\tt const} = {\tt polyConst(ℕ,𝔹)}.$$
Thus, adding polymorphism to the STLC has the effect of extending the notion
of type and term. With \(\lambda2\), we are now allowed to pass types as arguments
at term level (viz. we pass the actual types \(\mathbb{N}\) and \(\mathbb{B}\)).
It is, however, necessary to constrain this expressivity with a more expressive
type (e.g. {\tt (x : A) → ...}). Types that appear in this fashion are
called \(\Pi\)-types. Taking the grammar for simple types, we can add \(\Pi\)-types,
resulting in the grammar for types ({\tt t}) in \(\lambda2\):

\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
t || t → t || ∀ (x : Set) → t
\end{code}

@@

\begin{center}
\begin{minipage}{0.65\textwidth}
\begin{code}
t || t → t || ∀ (x : Set) → t || Set
\end{code}
\end{minipage}
\end{center}

The curious effect of doing this allows us to construct expressions of type
{\tt t → t}, where either {\tt t} can have subexpressions containing {\tt Set}.
Let's take the familiar definition of the list data structure:

@@

We discuss {\tt List} more in depth in a later section. For now, we simply ask
{\em what is the type of {\tt List}?} To answer this question, we can simply
look at the first line of the definition above (viz. {\tt List (A : Set) : Set}).
From this line, we can see that {\tt List} is a type parameterized over any type
{\tt A}. Another way to think about this statement is to think of {\tt List} as
a form of polymorphic function of type {\tt Set → Set}, as is similar with other
{\em parameterized types}. Unlike like {\tt const} or {\tt polyConst}, this
function can only be applied at type level (thus the name {\em type constructor}).
This is to insure that we cannot write type expressions that fail to instantiate
{\tt List} (e.g. {\tt List → A} v. {\tt List A → A}).

From this we can see that the addition of {\tt Set} into our grammar of
types has great effect on the expressivity of our system. There is only one
other extension necessary to consider our system fully-dependently typed, which,
similar to our treatment of \(\lambda\underline{\omega}\) is also deceptively
simple.

@@

\begin{center}
\begin{minipage}{0.7\textwidth}
\begin{code}
data _isEven : ℕ → Set where
  ZEven : 0 isEven
  NEven : ∀ n → n isEven → (Suc (Suc n)) isEven
\end{code}
\end{minipage}
\end{center}

Here, we've defined two ways in which {\tt n isEven} for all natural numbers
({\tt n}) can be proven. Returning to the idea of proof-relevant mathematics,
we do this by use of its two type constructors {\tt ZEven} and {\tt NEven},
corresponding to the two constructors of \(\mathbb{N}\). {\tt ZEven} is the
proof that 0 is an even number. {\tt NEven} then defines the property that if
a given natural number ({\tt n}) is even (viz. {\tt n isEven}), then {\tt n + 2}
is even (viz. {\tt (Suc (Suc n)) isEven}). To clarify how this type can be used,
let's prove that 4 is even. We start by providing the appropriate type definition
and also providing a name for our proof (viz. {\tt 4isEven}):
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
4isEven : 4 isEven
\end{code}
\end{minipage}
\end{center}
From the definition of {\tt isEven}, we know that to prove 4 is even, we must
know that 2 is even. Consequently, to prove 2 is even, we must know that 0 is
even. Fortunately, we have a proof that 0 is even: {\tt ZEven}. Thus, the proof
that 2 is even is {\tt NEven 0 ZEven}, and finally the proof that 4 is even is:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
4isEven : 4 isEven
4isEven = NEven 2 (NEven 0 ZEven)
\end{code}
\end{minipage}
\end{center}

@@

\subsection*{Calculus of Constructions and Remarks}
To conclude our discussion over dependent types, we present the final grammar for
types after extending \(\lambda\omega\) with \(\lambda P\):
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
t || t → t || ∀ (x : t) → t || Set
\end{code}
\end{minipage}
\end{center}
It should come with no surprise that the change is minor. We simply have lifted
the restriction on \(\Pi\)-types and allowed them to be parameterized over both
terms and types.

In the process of introducing dependent types, we have shown the necessary
changes and extensions on the STLC to achieve the expressivity necessary to
function as a foundation for mathematics. To clarify this process and to show
the relationship between each of the systems we have previously mentioned, we
include a diagram of the {\em Lambda Cube} \cite{}:
\begin{center}
  \begin{tikzpicture}[
      back line/.style={densely dotted},
      cross line/.style={preaction={draw=white, -,
          line width=6pt}}]
    \matrix (m) [matrix of math nodes,
      row sep=1em, column sep=.75em,
      text height=1ex,
      text depth=0.5ex]{
      & \lambda\omega & & \lambda C \\
      \lambda 2 & & \lambda P2 \\
      & \lambda\underline\omega & & \lambda P\underline\omega \\
      \lambda\!\!\rightarrow  & & \lambda P \\
    };
    \path[-]
    (m-1-2) edge (m-1-4)
    edge (m-2-1)
    edge [back line] (m-3-2)
    (m-1-4) edge (m-3-4)
    edge (m-2-3)
    (m-2-1) edge [cross line] (m-2-3)
    edge (m-4-1)
    (m-3-2) edge [back line] (m-3-4)
    edge [back line] (m-4-1)
    (m-4-1) edge (m-4-3)
    (m-3-4) edge (m-4-3)
    (m-2-3) edge [cross line] (m-4-3);
  \end{tikzpicture}
\end{center}
Reading the cube requires us to start on the edge marked \(\lambda\!\!\rightarrow\),
which corresponds to the STLC, the base system. From here, we can either move
inwards, upwards, or rightwards, which corresponds to the extension to the
other systems marked on the respective edges. Using only these three movements,
we eventually reach \(\lambda C\), leaving it in the {\em most powerful}
position on the cube.

@@

\subsection*{Principles}
Recursion and induction are one of the powerful tools available in a
fully-dependently typed system. Recursion and induction principles can be
thought of as specialized functions to behave on elements of a given type.
If we use recursion and induction principles to define functions and proofs,
we are guaranteed both termination and totality. This is because, in general,
recursion and induction principles are defined under certain known intuitions
(which is why they are sometimes considered axioms) on the given type that
they are meant to act on.

@@

\begin{minipage}{0.7\textwidth}
\begin{small}
The recursion and induction principles for some type {\tt T}.
\end{small}
\end{minipage}
\end{center}
The first argument given to these principles is sometimes called the {\em motive}.
This simply means that {\tt C} is the type from which we will return a element
from. For recursion principles, this means the given output of the function we
are defining, and for induction principles, this can be read as the proposition
we are trying to prove. The terms that follow (viz. {\tt C₀}) correspond to the
constructor(s) of the given type, which is how these principles guarantee
totality. Finally, both principles take a term of the type that are acting on
(viz. {\tt T} and {\tt (t : T)}). The only difference is that the induction
principle introduces a type family, as opposed to a non-parameterized type in the
recursion principle. We discuss this in Section 4.

@@

prop₀ : {A : Set} → ¬ (¬ (¬ A)) → ¬ A
prop₀ ¬¬¬a a = ?
\end{code}
\end{minipage}
\end{center}
Now, we solve for {\tt ?}. Looking at the type expression of {\tt prop₀},
we know we have an expression of type \(\neg\neg\neg A\) and also an
expression of type \(A\). To clarify, we can rewrite the type expression
after desugaring \(\neg A\):
$$\neg\neg\neg A\rightarrow A\rightarrow\perp$$
Now, to construct an element of type \(\neg A\) from an element of type
\(\neg\neg\neg A\), we can use the definition of \(\neg\):
$$(\neg\neg A\rightarrow\perp)\rightarrow A\rightarrow\perp$$
$$((\neg A\rightarrow\perp)\rightarrow\perp)\rightarrow A\rightarrow\perp$$
$$(((A\rightarrow\perp)\rightarrow\perp)\rightarrow\perp)\rightarrow A\rightarrow\perp$$
Thus, we know we must apply our expression \(\neg\neg\neg A\) to a function
of type \(\neg\neg A\) and so on, such that:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
prop₀ : {A : Set} → ¬ (¬ (¬ A)) → ¬ A
prop₀ ¬¬¬a a = ¬¬¬a (λ ¬a → ¬a a)
\end{code}
\end{minipage}
\end{center}
Or equivalently:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
prop₀ ¬¬¬a a = rec⊥ ⊥ (¬¬¬a (λ ¬a → ¬a a))
\end{code}
\end{minipage}
\end{center}

Next, we introduce the type \(\top\), which contains exactly one element,
{\tt *}:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data ⊤ : Set where
  * : ⊤

@@

ind⊤ : (C : ⊤ → Set) → C * → (x : ⊤) → C x
ind⊤ C c * = c
\end{code}
\end{minipage}
\end{center}
Similar to \(\perp\), \(\top\) is simple and is rather uninteresting by itself.
There are, however, some trivial yet interesting proofs that involve the two.
The first of these we also introduce a special kind of proof: {\it uniqueness
of identity} (UIP) \cite{}. In this proof, we set out to determine and classify
the identity of every element within a given type. For example, for \(\top\), we
should be able to prove:
$$\forall (x : \top) \rightarrow x ≡ *$$
That is, for every element of type \(\top\), that element is equivalent to {\tt *}.
This second expression, {\it x ≡} {\tt *}, is an example of an {\it identity type},
which we describe later on in this section. For now, we simply say that the
sole constructor for the identity type is {\it refl x}, for all x.

Now, to prove the above proposition, we can use the induction principle for
\(\top\), since the proposition we are trying to prove concerns all elements
of \(\top\):
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
uip⊤ : (x : ⊤) → x ≡ *
uip⊤ x = ind⊤ (λ top → top ≡ *) (refl *) x
--uip⊤ * = refl *
\end{code}
\end{minipage}
\end{center}
Here, we have taken our motive to be the expression {\tt (λ top → top ≡ *)},
which essentially repeats the information already given by our type expression.
The second element, {\tt refl *}, is a proof that our motive holds in the event
{\tt top} is {\tt *}, which is of type {\tt * ≡ *}. The last element {\tt x} is
the expression initially passed to our proof of type \(\top\). Equivalently, we
could have also written our proof as the third line above. This form of proof
uses pattern matching and essentially asserts that the only element in \(\top\)
is {\tt *}, which greatly simplifies our proof. Both methods of proof are
correct, but, in general, using pattern matching is only safe when the values of
proof expressions are not dependent one another (i.e., their values don't change
depending on which expression we apply pattern matching to).

@@

\subsection*{Extension to Homotopy Type Theory}

>>>>

\section{Introduction}
We present an introduction to dependent types and (Martin-L\"of) type theory.
At its core, type theory seeks to serve as a possible foundation of mathematics,
one that is based on a different logical foundation. This proof-relevant
form of mathematics is the basis in which types can be used to model logical
propositions and is the core idea of type theory. As such, not only does type
theory have a viable application in mathematics but also a growing interest
in the field of computer science. In both fields, type theory provides a richer
world of types and expressibiilty.

The study of type theory goes hand in hand with the study of dependent types,
which is why we first take the time to do familiarize ourselves with them
(Section 2). We then continue with the foundational concepts of type theory of
Per Martin-L\"of \cite{}. For this, we recommend using a proof assistant, such
as Agda. In our examples, we refer to Agda syntax rules.

@@

\begin{enumerate}
\item terms depending on terms
\item terms depending on types
\item types depending on types
\item types depending on terms
\end{enumerate}
These dependecies, however, should not be confused with dependent types.
In fact, only (2) and (4) correspond to expressions that are typed (necessarily)
with a dependent type (we explain later why (1) and (3) are not typed with
dependent types).

@@

\subsection*{Simply-Typed (STLC)}
\subsubsection*{Role: Base System}
The STLC is not a dependently typed system. In fact, the term {\em dependent type}
is drawn in constrast with the term {\em simple type}, where the STLC derives
its namesake. A simple type ({\tt t}) is either a {\em primitive type}, {\tt Prim},
which is the set of all types ``pre-defined'' (e.g. lists, numbers, booleans),
or a {\em function type} between other simple types. Thus, simple types have the
following grammar:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
Prim || t → t
\end{code}
\end{minipage}
\end{center}
The terms of the STLC are the terms of the \textit{untyped \(\lambda\)-calculus},
with the exception of terms that use \textit{self-application} (i.e., Y
combinator, Omega, etc.). An example term, annotated with its type, is generally
written as follows:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
example : ℕ → ℕ
example = λ (n : ℕ) → n
\end{code}
\end{minipage}
\end{center}
Or, equivalently,
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
example = λ n → n
\end{code}
\end{minipage}
\end{center}
whenever the type of the argument (viz. {\tt n}) is easily inferable. In this
case, since {\tt example} is of type {\tt ℕ → ℕ}, the argument {\tt n} has
to be of type {\tt ℕ}, therefore leaving its type annotation unnecessary.

The STLC has limited expressibilty but has the benefit of removing all
\textit{infinite} calculations from the system \cite{} and is thus the
appropriate foundation for deriving a type theoretic system. This is because
the STLC lends itself quite seamlessly as component of a larger, more expressive
system. This flexibility is what we exploit to derive \(\lambda C\). To do this,
we first extend the STLC with more expressive types through simple extentions on
its grammar for simple types.

The key difference between the STLC and \(\lambda C\) are the places in which
terms and types are allowed to be. In the STLC, terms and types are entirely
seperate entities in that terms only exist on {\em term level} and types
on {\em type level}. We return to the idea of {\em levels} later. Thus, as we
mentioned earlier, this puts a constraint on the system to only allow
\(\lambda\)-abstraction and function application to happen at term level
(viz. definition of simple type). For example, the \textit{constant function}
over types \(\mathbb{N}\) and \(\mathbb{B}\) in the STLC is defined:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
const : ℕ → 𝔹 → ℕ
const = λ n b → n
\end{code}
\end{minipage}
\end{center}
Given that {\tt const} is of type {\tt ℕ → 𝔹 → ℕ}, the term that inhabits this
type is {\tt λ n b → n}. In this example, we can see that terms and types are
treated differently and are only related in terms of {\em inhabitance}. This
relation is the reason for the limited expressivity of the STLC and is also the
method in which the STLC can be extended into \(\lambda C\).

@@

Another name for \(\lambda2\) is System F \cite{}. For our purposes, we treat
\(\lambda2\) simply as an extension of the STLC. This system is adds
{\em polymorphic} terms to the STLC (viz. Dependency (2)). For example,
let's take the definition of {\tt const} above and change it to define
{\tt polyConst}:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
polyConst : (A B : Set) → A → B → A
polyConst A B = λ a b → a
-- Works for any type A and B
\end{code}
\end{minipage}
\end{center}

Unlike {\tt const}, {\tt polyConst} takes four arguments, where the first two
are types. These two types (viz. {\tt A} and {\tt B}) are the corresponding
types of the third and fourth arguments of {\tt polyConst}. To clarify, we can
instantiate {\tt polyConst} with \(\mathbb{N}\) and \(\mathbb{B}\) resulting
in a function of type {\tt ℕ → 𝔹 → ℕ}, giving us:
$${\tt const} = {\tt polyConst(ℕ,𝔹)}.$$
Thus, adding polymorphism to the STLC has the effect of extending the notion
of type and term. With \(\lambda2\), we are now allowed to pass types as arguments
at term level (viz. we pass the actual types \(\mathbb{N}\) and \(\mathbb{B}\)).
It is, however, necessary to constrain this expressivity with a more expressive
type (e.g. {\tt (x : Set) → ...}). Types that appear in this fashion are
called \(\Pi\)-types. Taking the grammar for simple types, we can add \(\Pi\)-types,
resulting in the grammar for types ({\tt t}) in \(\lambda2\):

\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
Prim || t → t || ∀ (x : Set) → t
\end{code}

@@

\begin{center}
\begin{minipage}{0.65\textwidth}
\begin{code}
Prim || t → t || ∀ (x : Set) → t || Set
\end{code}
\end{minipage}
\end{center}
The curious effect of doing this allows us to construct expressions of type
{\tt t → t}, where either {\tt t} can have subexpressions containing {\tt Set}.
Let's take the familiar definition of the list data structure:

@@

We discuss {\tt List} more in depth in a later section. For now, we simply ask
{\em what is the type of {\tt List}?} To answer this question, we can simply
look at the first line of the definition above (viz. {\tt List (A : Set) : Set}).
From this line, we can see that {\tt List} is a type parameterized over any type
{\tt A}. Another way to think about this statement is to think of {\tt List} as
a form of polymorphic function of type {\tt Set → Set}, as is similar with other
{\em parameterized types}. Unlike like {\tt const} or {\tt polyConst}, this
function can only be applied at type level (thus the name {\em type constructor}).
This is to insure that we cannot write type expressions that fail to instantiate
{\tt List}. For example, the type expression {\tt List → A} is incorrect since
{\tt List} cannot appear in a type expression without first being instantiated.
To correct this, we can write this type expression as: {\tt List A → A}.

From this we can see that the addition of {\tt Set} into our grammar of
types has great effect on the expressivity of our system. There is only one
other extension necessary to consider our system fully-dependently typed, which,
similar to our treatment of \(\lambda\underline{\omega}\), is also deceptively
simple.

@@

\begin{center}
\begin{minipage}{0.7\textwidth}
\begin{code}
data _isEven : ℕ → Set where
  ZEven : 0 isEven
  NEven : ∀ n → n isEven → (Suc (Suc n)) isEven
\end{code}
\end{minipage}
\end{center}
Here, we've defined two ways in which {\tt n isEven} for all natural numbers
({\tt n}) can be proven. Returning to the idea of proof-relevant mathematics,
we do this by use of its two type constructors {\tt ZEven} and {\tt NEven},
corresponding to the two constructors of \(\mathbb{N}\). {\tt ZEven} is the
proof that 0 is an even number. {\tt NEven} then defines the property that if
a given natural number ({\tt n}) is even (viz. {\tt n isEven}), then {\tt n + 2}
is even (viz. {\tt (Suc (Suc n)) isEven}). To clarify how this type can be used,
let's prove that 4 is even. We start by providing the appropriate type definition
and also providing a name for our proof (viz. {\tt 4isEven}):
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
4isEven : 4 isEven
\end{code}
\end{minipage}
\end{center}
From the definition of {\tt isEven}, we know that to prove 4 is even, we must
know that 2 is even. Consequently, to prove 2 is even, we must know that 0 is
even. Fortunately, we have a proof that 0 is even: {\tt ZEven}. Thus, the proof
that 2 is even is {\tt NEven 0 ZEven}, and finally the proof that 4 is even is:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
4isEven : 4 isEven
4isEven = NEven 2 (NEven 0 ZEven)
\end{code}
\end{minipage}
\end{center}

@@

\subsection*{Calculus of Constructions and Remarks}
To conclude our discussion of dependent types, we present the final grammar for
types after extending \(\lambda\omega\) with \(\lambda P\), resulting in
\(\lambda C\):
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
Prim || t → t || ∀ (x : t) → t || Set
\end{code}
\end{minipage}
\end{center}
It should come with no surprise that the change is minor. We simply have lifted
the restriction on \(\Pi\)-types and allowed them to be parameterized over both
terms and types.

In the process of introducing dependent types, we have shown the necessary
changes and extensions on the STLC to achieve the expressivity necessary to
function as a foundation for mathematics. To clarify this process and to show
the relationship between each of the systems we have previously mentioned, we
include a diagram of the {\em Lambda Cube} \cite{}:
\begin{center}
  \begin{tikzpicture}[
      back line/.style={densely dotted},
      cross line/.style={preaction={draw=white, -,
          line width=6pt}}]
    \matrix (m) [matrix of math nodes,
      row sep=1em, column sep=.75em,
      text height=1ex,
      text depth=0.5ex]{
      & \lambda\omega & & \lambda C \\
      \lambda 2 & & \lambda P2 \\
      & \lambda\underline\omega & & \lambda P\underline\omega \\
      \lambda\!\!\rightarrow  & & \lambda P \\
    };
    \path[-]
    (m-1-2) edge (m-1-4)
    edge (m-2-1)
    edge [back line] (m-3-2)
    (m-1-4) edge (m-3-4)
    edge (m-2-3)
    (m-2-1) edge [cross line] (m-2-3)
    edge (m-4-1)
    (m-3-2) edge [back line] (m-3-4)
    edge [back line] (m-4-1)
    (m-4-1) edge (m-4-3)
    (m-3-4) edge (m-4-3)
    (m-2-3) edge [cross line] (m-4-3);
  \end{tikzpicture}
\end{center}
Reading the cube requires us to start on the edge marked \(\lambda\!\!\rightarrow\),
which corresponds to the STLC, the base system. From here, we can either move
inwards, upwards, or rightwards, which corresponds to the extension to the other
systems marked on the respective edges. Using these three movements, we
eventually reach \(\lambda C\), leaving it in the {\em most powerful} position
on the cube.

@@

\subsection*{Recursion and Induction Principles}
Recursion and induction are one of the powerful tools available in a
fully-dependently typed system. Recursion and induction principles can be
thought of as specialized functions to behave on elements of a given type.
If we use recursion and induction principles to define functions and proofs,
we are guaranteed both termination and totality. This is because, in general,
recursion and induction principles are defined under certain known intuitions
(which is why they are sometimes considered axioms) on the given type that
they are meant to act on.

@@

\begin{minipage}{0.7\textwidth}
\begin{small}
The recursion and induction principles for some type {\tt T}.
\end{small}
\end{minipage}
\end{center}
The first argument given to these principles is sometimes called the {\em motive}.
This simply means that {\tt C} is the type from which we will return an element
from. For recursion principles, this means the given output of the function we
are defining, and for induction principles, this can be read as the proposition
we are trying to prove. The terms that follow (viz. {\tt C₀}) correspond to the
constructor(s) of the given type, which is how these principles guarantee
totality. Finally, both principles take a term of the type that are acting on
(viz. {\tt T} and {\tt (t : T)}). The only difference is that the induction
principle introduces a type family, as opposed to a non-parameterized type in the
recursion principle. We discuss this in Section 4.

@@

prop₀ : {A : Set} → ¬ (¬ (¬ A)) → ¬ A
prop₀ ¬¬¬a a = ?
\end{code}
\end{minipage}
\end{center}
Now, we solve for {\tt ?}. Looking at the type expression of {\tt prop₀},
we know we have an expression of type \(\neg\neg\neg A\) and also an
expression of type \(A\). To clarify, we can rewrite the type expression
after desugaring \(\neg A\):
$$\neg\neg\neg A\rightarrow A\rightarrow\perp$$
Now, to construct an element of type \(\neg A\) from an element of type
\(\neg\neg\neg A\), we can use the definition of \(\neg\):
$$(\neg\neg A\rightarrow\perp)\rightarrow A\rightarrow\perp$$
$$((\neg A\rightarrow\perp)\rightarrow\perp)\rightarrow A\rightarrow\perp$$
$$(((A\rightarrow\perp)\rightarrow\perp)\rightarrow\perp)\rightarrow A\rightarrow\perp$$
Thus, we know we must apply our expression \(\neg\neg\neg A\) to a function
of type \(\neg\neg A\) and so on, such that:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
prop₀ : {A : Set} → ¬ (¬ (¬ A)) → ¬ A
prop₀ ¬¬¬a a = ¬¬¬a (λ ¬a → ¬a a)
\end{code}
\end{minipage}
\end{center}
Or equivalently:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
prop₀ ¬¬¬a a = rec⊥ ⊥ (¬¬¬a (λ ¬a → ¬a a))
\end{code}
\end{minipage}
\end{center}
Next, we introduce the type \(\top\), which contains exactly one element,
{\tt *}:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data ⊤ : Set where
  * : ⊤

@@

ind⊤ : (C : ⊤ → Set) → C * → (x : ⊤) → C x
ind⊤ C c * = c
\end{code}
\end{minipage}
\end{center}
Similar to \(\perp\), \(\top\) is simple and is rather uninteresting by itself.
There are, however, some trivial yet interesting proofs that involve the two.
For the first of these, we also introduce a special kind of proof: {\it uniqueness
of identity} (UIP) \cite{}. In this proof, we set out to determine and classify
the identity of every element within a given type. For example, for \(\top\), we
should be able to prove:
$$\forall (x : \top) \rightarrow x ≡ *$$
That is, for every element of type \(\top\), that element is equivalent to {\tt *}.
This second expression, {\it x ≡} {\tt *}, is an example of an {\it identity type},
which we describe later on in this section. For now, we simply say that the
sole constructor for the identity type is {\it refl x}, for all x.

Now, to prove the above proposition, we can use the induction principle for
\(\top\), since the proposition we are trying to prove concerns all elements
of \(\top\):
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
uip⊤ : (x : ⊤) → x ≡ *
uip⊤ = ind⊤ (λ top → top ≡ *) (refl *)
\end{code}
\end{minipage}
\end{center}
Here, we have taken our motive to be the expression {\tt (λ top → top ≡ *)},
which essentially repeats the information already given by our type expression.
The second element, {\tt refl *}, is a proof that our motive holds in the case that
{\tt top} is {\tt *} and is of type {\tt * ≡ *}.

Alternatively, we could have written our proof as follows:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
uip⊤ * = refl *
\end{code}
\end{minipage}
\end{center}
This form of proof uses pattern matching and essentially asserts that the only
element in \(\top\) is {\tt *}, which greatly simplifies our proof. Both methods
of proof are correct, but, in general, using pattern matching is only safe when
the values of proof expressions are not dependent one another (i.e., their
values don't change depending on which expression we apply pattern matching to).

@@

\subsection*{Extension to Homotopy Type Theory (maybe)}
